\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[round]{natbib}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue!60!black,citecolor=blue!60!black,urlcolor=blue!60!black]{hyperref}
\usepackage{microtype}
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

\newcommand{\R}{\mathbb{R}}
\newcommand{\silu}{\mathrm{SiLU}}
\newcommand{\T}{^{\top}}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\diag}{diag}

\title{How Much Tensor Structure Should a Transformer FFN Have?\\
\large Routed Block-CP Feedforward Networks Between SwiGLU and Dense Tucker}

\author{Aniket Deshpande\\
University of Illinois Urbana-Champaign\\
\texttt{aniket4@illinois.edu}}

\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
% TO FILL after results. Template (one claim per sentence):
% 1. obstacle: FFNs hold most params; multiplicative interactions hard to interpret.
% 2. exact observation: SwiGLU admits an exact routed CP decomposition.
% 3. the tension: superdiagonal core = atomized but one route per rank-1 atom.
% 4. proposal: LL1/block-CP family, per-route rank L as the dial.
% 5. experiments + main result.
% 6. careful implication.
[Abstract placeholder.]
\end{abstract}

\section{Introduction}
% P1: FFNs as the underinterpreted part of transformers.
% P2: exact tensor view of SwiGLU (words first).
% P3: CP's benefit and cost.
% P4: dense Tucker is the wrong naive move (honest revision).
% P5: LL1/block-CP structured relaxation; the gate-vs-atom budget trade.
% P6: empirical program.
% P7: contributions (numbered, true).
[Introduction placeholder.]

\section{The routed tensor view of gated FFNs}
\label{sec:theory}

\paragraph{Setup.}
At a fixed token position, let $x \in \R^d$ be the residual-stream vector entering the
feedforward block. The SwiGLU FFN \citep{shazeer2020glu} computes
\begin{equation}
\label{eq:swiglu}
y = U\T h, \qquad h = (W\T x) \odot \silu(G\T x),
\end{equation}
with $W, G \in \R^{d\times m}$ and $U \in \R^{m \times d}$. Writing $w_j, g_j$ for the
$j$-th columns of $W, G$ and $u_j$ for the $j$-th column of $U\T$, the $j$-th hidden
coordinate is $h_j(x) = (w_j\T x)\,\silu(g_j\T x)$.

\subsection{SwiGLU is an exactly routed CP tensor field}
\label{sec:routedcp}

Each SwiGLU channel multiplies two scalar projections of the residual stream. The SiLU
nonlinearity factorizes exactly, $\silu(z) = z\,\sigma(z)$ with
$\sigma(z) = (1+e^{-z})^{-1}$, so
\begin{equation}
\label{eq:factorized}
h_j(x) \;=\; \underbrace{(w_j\T x)(g_j\T x)}_{\text{rank-one bilinear interaction}}
\cdot \underbrace{\sigma(g_j\T x)}_{\alpha_j(x)\,\in\,(0,1)} .
\end{equation}
Defining the input-dependent interaction tensor $A(x) \in \R^{d\times d\times d}$ by
$y_o = \sum_{a,b} A_{oab}(x)\, x_a x_b$, substituting \eqref{eq:factorized} into
\eqref{eq:swiglu} gives
\begin{equation}
\label{eq:cp}
A(x) \;=\; \sum_{j=1}^{m} \alpha_j(x)\; u_j \otimes w_j \otimes g_j ,
\qquad \alpha_j(x) = \sigma(g_j\T x).
\end{equation}
This is a canonical polyadic (CP) decomposition \citep{kolda2009tensor} of $A(x)$ with
\emph{shared, input-independent factors} and \emph{input-dependent routing weights}
$\alpha_j(x)$. The decomposition is exact and holds pointwise in $x$; no approximation
is involved. We call $u_j \otimes w_j \otimes g_j$ the $j$-th \emph{interaction atom}
(what interaction is available) and $\alpha_j(x)$ its \emph{route} (when it is used).
The bilinear $\alpha\equiv 1$ limit of this picture is the bilinear MLP analyzed by
\citet{pearce2025bilinear}; the routed form keeps the gate inside the structural
description rather than discarding it.

Viewed as a Tucker decomposition with latent dimensions $r=s=m$ and factor matrices
$(P,Q,R) = (W, G, U\T)$, the CP form \eqref{eq:cp} corresponds to a
\emph{superdiagonal core}: $C_{oij} = \delta_{oi}\delta_{ij}$. This restriction cuts
both ways. It atomizes the layer --- every hidden index is one (atom, route) pair, and
CP decompositions are essentially unique under mild conditions
\citep{kruskal1977three}, so the atoms are identifiable objects. But it also means one
route controls exactly one rank-one interaction: feature $w_i$ never interacts with
gate $g_j$ for $i \ne j$.

\subsection{Dense Tucker and the per-route interaction matrix}
\label{sec:tucker}

The unconstrained relaxation drops the superdiagonal restriction. A
\emph{Tucker-core FFN} with latent projections $P, Q \in \R^{d\times r}$, output basis
$R \in \R^{d \times s}$, and learned core $C \in \R^{s\times r\times r}$ computes
$p = P\T x$, $q = Q\T x$,
\begin{equation}
\label{eq:tuckerffn}
z_o = \sum_{i,j=1}^{r} C_{oij}\, p_i\, \silu(q_j), \qquad y = Rz .
\end{equation}
Grouping by gate index exposes the structure this paper is about:
\begin{equation}
\label{eq:vj}
y(x) = \sum_{j=1}^{r} V_j\, p \; \silu(q_j), \qquad V_j := R\,C^{(j)} \in \R^{d\times r},
\quad C^{(j)}_{oi} = C_{oij}.
\end{equation}
Each latent gate $q_j$ routes a \emph{matrix} $V_j$ of output-by-main interactions.
SwiGLU is the corner case $\rank V_j = 1$ for all $j$; a generic Tucker core has
$\rank V_j = \min(r,s)$.

Two properties make the dense core suspicious as an interpretability target. First,
\emph{gauge freedom}: for any invertible $M \in \R^{r\times r}$, replacing $P \mapsto
P M^{-\top}$ and $C^{(j)} \mapsto C^{(j)} M\T$ for all $j$ leaves the function
unchanged, so individual latent directions $p_i$ and individual core entries carry no
intrinsic meaning --- only the gate directions $Q_{:j}$ and the subspaces
$\mathrm{range}(V_j)$ are well defined. Second, the core costs $s r^2$ parameters; at
the configurations we study it absorbs over $90\%$ of the block's budget.

\paragraph{The aligned-width theorem, reinterpreted.}
Prior work on this codebase proved an exact separation: if $[P\;Q]$ has full column
rank, an \emph{aligned} SwiGLU (units restricted to main vectors in $\mathrm{span}(P)$
and gate vectors from the columns of $Q$) that exactly realizes \eqref{eq:vj} must have
width $m \ge \sum_j \rank V_j$, and a rank decomposition of each $V_j$ attains the
bound (Appendix~\ref{app:theorem}). Generically this is $r\min(r,s)$, i.e.\ $k^2$ in
the balanced case --- the result that originally motivated dense Tucker as
``$k\times$ cheaper than SwiGLU.'' The reading we adopt here is different: \emph{the
theorem says the operative variable is the per-route rank $\rank V_j$}, not the
presence of a dense core. The constructive half of the proof --- realize each $V_j$ as
a sum of $\rank V_j$ rank-one terms sharing the same gate --- is itself an
architecture. That architecture is the subject of this paper.

\subsection{LL1/block-CP: one route, one rank-$L$ block}
\label{sec:ll1}

A third-order tensor admits a \emph{decomposition in multilinear rank-$(L,L,1)$ terms}
(LL1; a structured block-term decomposition) if it can be written
$T = \sum_{b=1}^{B} (A_b B_b\T) \otimes c_b$: each term is rank-$L$ in two modes and
rank-one in the third \citep{delathauwer2008blockterm,vervliet2016tensorlab}. Taking
the \emph{gate mode} as the rank-one mode yields the FFN we study. The
\textbf{LL1 FFN} with $B$ blocks of rank $L$ has, per block, a gate direction
$g_b \in \R^d$, a main factor $A_b \in \R^{d\times L}$, and an output factor
$U_b \in \R^{d \times L}$:
\begin{equation}
\label{eq:ll1}
y(x) \;=\; \sum_{b=1}^{B} U_b \big(A_b\T x\big)\, \silu(g_b\T x).
\end{equation}
Expanding into atoms, $y(x) = \sum_b \sum_{l=1}^{L} u_{b,l} (a_{b,l}\T x)\,
\silu(g_b\T x)$: \emph{grouped CP}, $L$ rank-one atoms sharing one route. The
interaction tensor is the routed LL1 tensor
$A(x) = \sum_b \alpha_b(x)\, (U_b A_b\T) \otimes g_b$, and the per-route matrix is
$V_b = U_b A_b\T$ with $\rank V_b \le L$ \emph{by construction}. The aligned-width
theorem's control variable has become a hyperparameter.

The family nests both endpoints exactly. At $L=1$, $B=m$, \eqref{eq:ll1} \emph{is}
SwiGLU (verified to machine precision in our implementation); a single block with
$L = r$ is a (one-gate) dense slice; and an LL1 FFN equals a Tucker FFN whose core is
block-superdiagonal. LL1 decompositions also retain essential-uniqueness guarantees
under mild conditions \citep{delathauwer2008blockterm,delathauwer2011blockterm},
unlike the gauge-free dense core.

\paragraph{What the parameter budget buys.}
The comparison that matters is at matched parameters:
\begin{equation}
\label{eq:params}
N_{\text{SwiGLU}} = 3dm, \qquad
N_{\text{LL1}} = dB(2L+1), \qquad
N_{\text{Tucker}} = d(2r+s) + sr^2 .
\end{equation}
At a fixed budget $N$, SwiGLU affords $m = N/3d$ atoms, each with a private route; LL1
affords $BL = \tfrac{3L}{2L+1}\cdot\tfrac{N}{3d}$ atoms --- up to $1.5\times$ more ---
but only $B = \tfrac{3}{2L+1}\cdot\tfrac{N}{3d}$ routes. \emph{LL1 trades routing
diversity for interaction atoms.} An equivalent statement: an LL1$(B,L)$ block is
exactly a width-$BL$ SwiGLU whose gate vectors are tied in groups of $L$. Which side
of the trade wins is an empirical question about what trained FFNs actually need, and
it is the question our experiments are designed to answer. A prior observation
sharpens the stakes: in a dense Tucker LM trained from scratch, the learned per-gate
stable rank $\|V_j\|_F^2/\|V_j\|_{\mathrm{op}}^2$ concentrates around $4$ --- far
below the generic value $r$, far above the SwiGLU value $1$. The trained dense core
already behaves like an LL1 model; LL1 makes that behavior an explicit, cheaper, and
identifiable parameterization.

% TABLE: matched configs (B, L, atoms, routes) at d=512 budget 2.293M.

\section{Experiments}
\label{sec:experiments}
% Each subsection: Question / Setup / Metric / Result / Interpretation / Caveat.

\subsection{Synthetic teacher--student recovery}
\label{sec:synthetic}
[exp18 placeholder]

\subsection{Matched-budget language modeling}
\label{sec:lm}
[exp11/sprint\_lm placeholder]

\subsection{Interpretability proxies}
\label{sec:interp}
[exp19/exp22 placeholder]

\subsection{Circuit-level pilot: induction}
\label{sec:induction}
[exp20 placeholder]

\section{Discussion}
\label{sec:discussion}
[placeholder: what is established / plausible / open; limitations; next experiment.]

\section{Related work}
\label{sec:related}

\paragraph{Gated FFNs and bilinear MLPs.}
GLU-family FFNs outperform ReLU/GELU MLPs at matched scale and are standard in modern
LMs \citep{shazeer2020glu}. \citet{pearce2025bilinear} study bilinear MLPs --- GLUs
with the element-wise nonlinearity removed --- which are exactly third-order tensors,
and show that eigendecomposition of those tensors yields interpretable low-rank
structure; \citet{dooms2024case} make the architectural case. Our object differs in
keeping the gate: SwiGLU is exactly a \emph{routed} CP tensor field whose routing is
input-dependent, and prior experiments on this codebase show the routing is
load-bearing (constant-$\alpha$ ablations degrade pretrained Qwen models by four to
six orders of magnitude in perplexity), so the bilinear picture alone does not
describe trained SwiGLU models. \citet{jayakumar2020multiplicative} situate gating
within multiplicative interactions broadly; we study the specific tensor
decompositions a gated FFN admits.

\paragraph{Tensor decompositions in neural networks.}
CP, Tucker, and block-term decompositions are classical
\citep{kolda2009tensor,delathauwer2008blockterm}. In deep learning they are most
often used for \emph{compression} of trained weights \citep{novikov2015tensorizing}
or for expressivity analyses of idealized architectures \citep{cohen2016expressive};
factorization machines use low-rank multiplicative atoms for feature interactions
\citep{rendle2010factorization}; \citet{loconte2024relationship} connect tensor
factorizations to probabilistic circuits. Our use is different: the decomposition is
not applied post hoc to a weight tensor --- the FFN \emph{is} the decomposition, and
we ask which member of the CP$\to$LL1$\to$Tucker family is the right parameterization
to train.

\paragraph{Sparse dictionaries and weight-based interpretability.}
Sparse autoencoders and transcoders decompose activations into feature directions
\citep{bricken2023monosemanticity,dunefsky2024transcoders,paulo2025transcoders}.
The routed-CP/LL1 view is complementary: atoms and routes are \emph{architectural}
objects fixed in the weights, with the input-dependence isolated in scalar routing
coefficients --- the same invariant-dictionary + input-dependent-coefficients
factorization transcoders recover empirically.

\paragraph{Structured matrices for efficiency.}
Monarch and butterfly factorizations replace dense projections with products of
block-diagonal or sparse factors \citep{dao2022monarch,dao2019butterfly}; LoRA
demonstrates that rank constraints are effective in transformers \citep{hu2022lora}.
These address the \emph{factor matrices}; our axis is the \emph{latent core}
connecting them. The two are orthogonal and composable.

\paragraph{Circuit-level evaluation.}
Induction heads are the canonical recognizable circuit
\citep{elhage2021mathematical,olsson2022context}; we use a minimal induction task to
ask whether FFN tensor structure changes the emergence or mechanism of an
attention-side circuit.

\bibliographystyle{plainnat}
\bibliography{refs}

\appendix

\section{The aligned-width theorem}
\label{app:theorem}

This appendix restates, with proof, the separation theorem from the earlier draft
accompanying this codebase, and records its constructive half as the LL1
architecture. Throughout, $\phi = \silu$; the only properties used are $\phi(0)=0$
and $\phi(1) \ne 0$.

\begin{lemma}[SwiGLU recovery]
\label{lem:recovery}
Setting $r=s=m$, $P=W$, $Q=G$, $R=U\T$, and $C_{oij} = \delta_{oi}\delta_{ij}$ in
\eqref{eq:tuckerffn} reduces the Tucker-core FFN to the SwiGLU map \eqref{eq:swiglu}.
\end{lemma}

\begin{proof}
With the superdiagonal core, \eqref{eq:tuckerffn} collapses to
$z_o = p_o\,\phi(q_o)$, so $z = (P\T x) \odot \phi(Q\T x)$, which is $h$ from
\eqref{eq:swiglu}; then $y = Rz = U\T h$.
\end{proof}

\begin{definition}[Aligned SwiGLU representation]
\label{def:aligned}
Given latent matrices $P, Q \in \R^{d\times r}$, an \emph{aligned width-$m$ SwiGLU
representation} is a map
\[
\hat y(x) = \sum_{j=1}^{r} \sum_{\nu=1}^{k_j} u_{j\nu}\, (a_{j\nu}\T p)\,\phi(q_j),
\qquad \textstyle\sum_j k_j = m,
\]
with $u_{j\nu} \in \R^d$, $a_{j\nu} \in \R^r$, $p = P\T x$, $q = Q\T x$. Each unit
uses a main vector in $\mathrm{span}(P)$ and one of the gate vectors $Q_{:j}$.
\end{definition}

\begin{theorem}[Diagonal bottleneck]
\label{thm:bottleneck}
Assume $[P\;Q] \in \R^{d\times 2r}$ has full column rank. The minimum width of an
aligned SwiGLU representation exactly realizing a Tucker-core FFN with per-gate
coefficient matrices $\{V_j\}_{j=1}^r$ is $m_{\min} = \sum_{j=1}^{r} \rank V_j$.
\end{theorem}

\begin{proof}
\emph{Lower bound.} By the rank assumption, $x \mapsto (p, q)$ is surjective onto
$\R^r \times \R^r$, so an exact representation forces
$\sum_j (V_j - \hat V_j)\, p\, \phi(q_j) = 0$ for all $(p,q)$, where
$\hat V_j = \sum_\nu u_{j\nu} a_{j\nu}\T$ has rank at most $k_j$. Fix $j_0$ and set
$q_j = \delta_{jj_0}$; since $\phi(0) = 0$ and $\phi(1) \ne 0$, only the $j_0$ term
survives and $V_{j_0} = \hat V_{j_0}$. Hence $k_j \ge \rank V_j$ for every $j$, and
$m = \sum_j k_j \ge \sum_j \rank V_j$.

\emph{Tightness.} Let $V_j = \sum_{\nu=1}^{\rho_j} u_{j\nu} a_{j\nu}\T$ be any rank
decomposition, $\rho_j = \rank V_j$. Then
$V_j\, p\, \phi(q_j) = \sum_\nu u_{j\nu} \big((P a_{j\nu})\T x\big)\,
\phi(Q_{:j}\T x)$ is a sum of $\rho_j$ aligned units sharing gate $Q_{:j}$.
\end{proof}

\begin{corollary}[Generic separation]
If $R$ has full column rank and each slice $C^{(j)}$ is generic, then
$\rank V_j = \min(r,s)$ for all $j$ and $m_{\min} = r \min(r,s)$; in the balanced
case $r=s=k$, $m_{\min} = k^2$.
\end{corollary}

\begin{remark}[The constructive half is LL1]
The tightness construction --- $\rho_j$ rank-one units sharing the gate $Q_{:j}$ ---
is precisely one block of the LL1 FFN \eqref{eq:ll1} with $L = \rho_j$. Theorem
\ref{thm:bottleneck} therefore does not say ``dense cores beat SwiGLU''; it says the
cost of simulating a routed tensor map is the sum of its per-route ranks, and an
architecture that fixes those ranks directly realizes the map at exactly that cost.
\end{remark}

\section{Experimental details}
\label{app:details}
[hyperparameters, configs, hardware.]

\end{document}
